Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-06-24 paper/paper.Rmd
add intro to "clustering" and "spatial" chapter, adjust heading levels, change titles

\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
Several of the implemented algorithms are themselves versatile toolboxes with multiple options. Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>). The following sections give an overview of most existing partitioning strategies and their visualization. The available methods are further discussed in section \@ref(sec:disc).

\begin{table}[h]
  \centering
  \footnotesize
  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (Notation column). Column count refers to the number of studies which were found to have used this resampling technique until 2021-05.}
  \begingroup
    \begin{tabular}{lllll}
      \\
      Type                            & Name                        & R Package(s)     & mlr3 Notation  & Count\\
      \toprule
      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}}  \tablefootnote{\label{diesing2020}\cite{diesing2020}} \\  \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}}}  & \specialcell{2 \\ 3}   \\
      \cmidrule{2-5}
                                      & Spatial Blocking            & \specialcell{\pkg{blockCV}  \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} \\  \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} }
                                        & \specialcell{28 \\ 4} \\
  \cmidrule{2-5}
                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} &  21 \\
                                      \cmidrule{2-5}
       & Group-Level CV & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} (LLO) \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"custom\_cv"} \tablefootnote{\label{brenning2006}\cite{brenning2006}} \tablefootnote{\label{pena2015}\cite{pena2015}}} & \specialcell{12 \\ 2}       \\
      \midrule
      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"} \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & 6 \\
                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} &  -  \\
      \midrule
      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"} \footref{morera2021}   & 1    \\

@@

@@

Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
\pkg{mlr3spatiotempcv} supports two "buffering" implementations: (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}.

@@

### Spatial Clustering (sperrorest)

#### Coordinate-based clustering

The "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
$k$-means clustering has been proposed as a clustering method.
To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}.

@@

## Spatiotemporal

@@

#### Custom (fixed) Resampling --- mlr3

@@

### Additional partitioning techniques

@@

<!-- missing: non-LOO method with buffer support -->
Also no methods yet exists that makes use of a buffer zone with blocking or clustering, bridging the gap between buffered LOO-CV and fold-based non-buffer methods (`"spcv_coords"` and `"spcv_block"`).

>>>>

\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
The following sections, which follow the naming scheme `<method name> --- <R package>`, give an overview of most existing partitioning strategies and their visualization.
The available methods are further discussed in section \@ref(sec:disc).

\begin{table}[h]
  \centering
  \footnotesize
  \caption[t]{Spatiotemporal resampling methods, ordered alphabetically by class name (column 'mlr3 Notation'). Column count refers to the number of studies which were found to have used this resampling technique until 2021-05.}
  \begingroup
    \begin{tabular}{lllll}
      \\
      Type                            & Name                        & R Package(s)     & mlr3 Notation  & Count\\
      \toprule
      \multirow{6}{*}{Spatial}        & Spatial Buffering           & \specialcell{\pkg{blockCV} \\ \pkg{sperrorest}}    & \specialcell{ \texttt{"spcv\_buffer"} \tablefootnote{\label{ploton2020}\cite{ploton2020}}  \tablefootnote{\label{diesing2020}\cite{diesing2020}} \\  \texttt{"spcv\_disc"} \tablefootnote{\label{endicott2017}\cite{endicott2017}} \tablefootnote{\label{karasiak2021}\cite{karasiak2021}} \tablefootnote{\label{moller2021}\cite{moller2021}}}  & \specialcell{2 \\ 3}   \\
      \cmidrule{2-5}
                                      & Spatial Blocking            & \specialcell{\pkg{blockCV}  \\ \pkg{sperrorest}}  & \specialcell{  \texttt{"spcv\_block"} \tablefootnote{\label{jensen2021}\cite{jensen2021}} \tablefootnote{\label{escobar2021}\cite{escobar2021}} \tablefootnote{\label{stewart2021}\cite{stewart2021}} \\  \texttt{"spcv\_tiles"} \tablefootnote{\label{bebber2017}\cite{bebber2017}} \tablefootnote{\label{brenning2015}\cite{brenning2015}} \tablefootnote{\label{zurell2020}\cite{zurell2020}} }
                                        & \specialcell{28 \\ 4} \\
  \cmidrule{2-5}
                                      & Spatial Clustering          & \pkg{sperrorest} & \texttt{"spcv\_coords"} \tablefootnote{\label{morera2021}\cite{morera2021}} \tablefootnote{\label{geiss2017}\cite{geiss2017}} \tablefootnote{\label{wu2020}\cite{wu2020}} &  21 \\
                                      \cmidrule{2-5}
       & Group-Level CV & \specialcell{\pkg{CAST} \\ \pkg{mlr3}}       & \specialcell{ \texttt{"sptcv\_cstf"} (LLO) \tablefootnote{\label{blothe2021}\cite{blothe2021}} \tablefootnote{\label{dolan2021}\cite{dolan2021}} \tablefootnote{\label{dumke2019}\cite{dumke2019}}  \\ \texttt{"custom\_cv"} \tablefootnote{\label{brenning2006}\cite{brenning2006}} \tablefootnote{\label{pena2015}\cite{pena2015}}} & \specialcell{12 \\ 2}       \\
      \midrule
      \multirow{2}{*}{Spatiotemporal} & Leave-Location-and-Time-Out & \pkg{CAST}       &  \texttt{"sptcv\_cstf"} \tablefootnote{\label{egli2020}\cite{egli2020}} \tablefootnote{\label{gao2019}\cite{gao2019}} \tablefootnote{\label{reitz2021}\cite{reitz2021}} & 6 \\
                                      & Spatiotemporal Clustering   & \pkg{skmeans}    &  \texttt{"sptcv\_cluto"} &  -  \\
      \midrule
      Feature space                   & Environmental Blocking      & \pkg{blockCV}    &  \texttt{"spcv\_env"} \footref{morera2021}   & 1    \\

@@

Spatial partitioning methods operate in a two-dimensional space and split observations into training and test set by their spatial location.
Methods such as grid-based partitioning or clustering are used to create groups of observations which may then be aggregated further to form the final partition sets.

Judging by their common characteristics and the available implementations, spatial partitioning methods can be coarsely categorized into (1) buffering, (2) blocking and (3) clustering methods.
If a method had the potential to belong to multiple categories, a deliberate category assignment was made for one specific group which matched the main characteristic of the respective method best.

@@

Buffering is mostly applied in the coordinate space due to the scattered appearance of observations in this space.
However, there is no practical limitation that would prevent this method to be applied in a 2D + time setting.
Because \pkg{mlr3spatiotempcv} supports two spatial-only "buffering" implementations, (1) "Spatial buffering" from package \pkg{blockCV} and (2) "Leave-one-disc-out cross-validation" from \pkg{sperrorest}, buffering was attributed to the category of spatial partitioning methods.

@@

### Clustering

The idea of cluster analysis, which originated in the 1930s in the anthropology field [@driver1932, @zubin1938], is to group objects with similar characteristics.
In the spatial context, the similarity of spatial coordinates is used to form such groups.
Different cluster algorithms such as hierarchical, density-based, grid-based or centroid-based clustering exist [@gower1967].

#### Coordinate-based clustering --- sperrorest

The so-called "Spatial CV" method after [@russ2010; @brenning2012] and implemented in package \pkg{sperrorest} uses the coordinates of all observations to create clusters in the spatial domain.
$k$-means clustering, which belongs to the centroid-based clustering family, has been proposed as a clustering method.
To avoid confusion with other spatial partitioning methods, we refer to this as coordinate-based clustering, or as `spcv_coords` in \pkg{mlr3spatiotempcv}, instead of using the generic "Spatial CV" term.

@@

## Spatial or Spatiotemporal

<!-- FIXME: explain that most methdso can also be spatial-only -->

@@

### Custom (fixed) Resampling --- mlr3

@@

## Additional partitioning techniques

@@

<!-- FIXME: mention missing spatiotemporal buffering method -->

<!-- missing: non-LOO method with buffer support -->
<!-- FIXME: this is wrong, leave-one-disc-out does this -->
Also no methods yet exists that makes use of a buffer zone with blocking or clustering, bridging the gap between buffered LOO-CV and fold-based non-buffer methods (`"spcv_coords"` and `"spcv_block"`).
